---
title: "CS201 - Introduction & Getting Started with R & RStudio"
subtitle: "GitHub, warm-up, and a first look at mtcars"
author: "Amber Camp"
date: "2025-09-08"
format: html
editor: visual
---

## Welcome Back!

-   Quick check-in: Who still needs help getting R, RStudio, and Git installed?
-   Plan for today:
    -   Revisit: R vs. RStudio
    -   Git vs GitHub & why we care
    -   Base R vs. tidyverse (conceptual only)
    -   Play with `mtcars`

------------------------------------------------------------------------

## R vs. RStudio

What is the difference between R and RStudio and why do you need to have both for this class?

-   **R** is the programming language and engine
-   **RStudio** is the IDE (workspace) that makes using R easier
-   You *can* run R without RStudio, but you probably wouldn’t want to (ask me how I know)

------------------------------------------------------------------------

## Git vs. GitHub

Similarly, remember that Git and GitHub are *not the same thing*.

-   **Git**: Tracks changes to your code locally on your computer
-   **GitHub**: Hosts your Git repositories online for sharing, backup, and collaboration
-   Together: You can track your work + share it with others

------------------------------------------------------------------------

## Extra Credit Question

Who was able to figure out why `0.1 + 0.1 + 0.1 - 0.3 ≠ 0`?

Try it!

```{r}
0.1 + 0.1 + 0.1 - 0.3
0.3 - 0.1 - 0.1 - 0.1 # a variation

# run this if you want to avoid scientific notation
options(scipen = 999)
```

Explanation:

-   

------------------------------------------------------------------------

# Let's get started!

## Install & Load packages

This gives you access to packages/libraries, which are like custom/specialized tools so that you can do more with R.

```{r}
# install.packages("tidyverse")
library(tidyverse)
```

## Quick note on Base R vs. `tidyverse`

-   **Base R**: built-in functions, available immediately
-   **`tidyverse`**: collection of modern packages with consistent syntax
-   **`tidyverse`** is considered the modern way to interact with R, and this will be our focus

### Example

Both of the following create a table from the dataset `mtcars` with columns for `mpg` and `hp`, filtering for 6 cylinder models and `mpg` \> 20.

**Base R:**

```{r}
mtcars[mtcars$cyl == 6 & mtcars$mpg > 20, c("mpg", "hp")]
```

**tidyverse:**

```{r}
mtcars %>%
  filter(cyl == 6, mpg > 20) %>%
  select(mpg, hp)
```

-   Base R uses bracket notation and logical conditions
-   `tidyverse` uses verbs like (`filter()`, `select()`) chained with pipes for readability
-   Base R powerful and always available. `tidyverse` makes code more expressive and human-readable, especially for multi-step data wrangling (it's often easier to use!)

------------------------------------------------------------------------

## First Look: `mtcars`

`mtcars` is a classic, built-in dataset. If you've used R before, you've probably interacted with `mtcars`.

Start by exploring the data below. It's always good to ensure you understand what is contained in your dataset.

```{r}
data(mtcars) # reads in the data & adds it to your environment as a dataframe
View(mtcars) # opens up the datafram for viewing
head(mtcars) # displays the top few lines of the dataframe
summary(mtcars) # displays a summary of all variables
?mtcars # will open up documentation in your Help tab. works for packages, functions, too!
```

-   How many rows and columns are in the dataframe?
-   What variables look interesting?
-   How might you view the last few rows of the dataframe with a single function?

------------------------------------------------------------------------

### Quick Questions

-   What is the average `mpg` overall?

```{r}
mean(mtcars$mpg) # the $ helps you select a specific column of interest
```

-   Can you find the mean of another variable?

```{r}

```

-   Can you find the median?

```{r}

```

------------------------------------------------------------------------

### Quick Plot

We are using *Base R* here, and it works fine! Going forward, we'll use a packaged called `ggplot2`.

```{r}
# this is a scatterplot of mpg by wt
plot(mtcars$wt, mtcars$mpg,
     xlab = "Weight (1000 lbs)", ylab = "MPG",
     main = "Heavier cars tend to have lower MPG")
```

------------------------------------------------------------------------

## Exit Ticket

Create a scatterplot of two variables of your choice (but not `mpg` vs. `wt`). Feel free to try a `ggplot2` version if you have some R experience!

```{r}

```

Save, stage, commit, and push to GitHub!!
